---
title: "Figures"
author: "Skanderup lab"
date: "9/19/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(wesanderson)

tumor.colors = list()
tumor.colors['HNSC'] = '#26A9E0'
tumor.colors['GBM'] = '#6D6E70'
tumor.colors['LGG'] = '#A6A8AB'
tumor.colors['THCA'] = '#50A38C'
tumor.colors['SKCM'] = '#6BD9BA'
tumor.colors['BRCA'] = '#DB2F91'
tumor.colors['LUAD'] = '#F89E21'
tumor.colors['LUSC'] = '#FCC27B'
tumor.colors['ESCA'] = '#7A4F9F'
tumor.colors['STAD'] = '#A58AC0'
tumor.colors['CRC'] = '#DAD1E8'
tumor.colors['KIRP'] = '#C4E7ED'
tumor.colors['KIRC'] = '#72CAC8'
tumor.colors['LIHC'] = '#E2DC1B'
tumor.colors['BLCA'] = '#6EBD45'
tumor.colors['PAAD'] = '#B5D789'
tumor.colors['PRAD'] = '#2B388F'
tumor.colors['OV'] = '#EC2224'
tumor.colors['CESC'] = '#F48466'
tumor.colors['UCEC'] = '#FBD2C0'
tumor.colors = unlist(tumor.colors)

df = read.csv('data/pancancer_product_score.csv')
df$r = sub('.+_',"\\1",df$lr)
df$l = sub('_.+',"\\1",df$lr)
df$ss = 100 - df$cc - df$cs - df$sc - df$nn

expr = read.csv('data/deconv_logx+1.csv')
colnames(expr)[1] = 'gene'
rownames(expr) = expr[,1]

```


# Pan-cancer crosstalk analysis

## Expression of ligands and receptors across tumor types

```{r}

# pan-cancer expression of ligands
df.l = unique(df[,c('l','t','c_l','s_l')])
df.l = df.l[df.l$t!='MEDIAN',]
df.l$lfc = log2((df.l$c_l+1)/(df.l$s_l+1))
df.l = df.l[order(df.l$lfc),]
df.l.pancan = aggregate(df.l$lfc,list(df.l$l),median)
colnames(df.l.pancan) = c('l','lfc')
df.l.pancan = df.l.pancan[order(df.l.pancan$lfc),]

# pan-cancer expression of receptors
df.r = unique(df[,c('r','t','c_r','s_r')])
df.r = df.r[df.r$t!='MEDIAN',]
df.r$lfc = log2((df.r$c_r+1)/(df.r$s_r+1))
df.r = df.r[order(df.r$lfc),]
df.r.pancan = aggregate(df.r$lfc,list(df.r$r),median)
colnames(df.r.pancan) = c('r','lfc')
df.r.pancan = df.r.pancan[order(df.r.pancan$lfc),]

par(mfrow=c(1,2))
barplot(df.l.pancan$lfc,ylim=c(-7,2))
barplot(df.r.pancan$lfc,ylim=c(-7,2))


# list top-10 ligands and receptors
head(df.l.pancan,n=10)
tail(df.l.pancan,n=10)[10:1,]
head(df.r.pancan,n=10)
tail(df.r.pancan,n=10)[10:1,]

```

## Pan-cancer crosstalk scores

```{r}

# cc cs sc ss nn
ct.colors = cbPalette <- c("#D55E00", "#E69F00", "#56B4E9", "#0072B2",  "#BBBBBB")

df.med = list()
for(lr2 in unique(df$lr)) {
  xx = df %>% filter(lr == lr2)
  rr = c(median(xx$cc,na.rm = T),median(xx$ss,na.rm = T),median(xx$sc,na.rm = T),median(xx$cs,na.rm = T),median(xx$nn,na.rm = T))
  df.med[[lr2]] = rr
}
df.med = data.frame(do.call(rbind,df.med))
df.med$lr = rownames(df.med)
colnames(df.med) = c('cc','ss','sc','cs','nn','lr') 
df.med$r = sub('.+_',"\\1",df.med$lr)

df.med.g = gather(df.med,'type','value',-r,-lr,-nn)
p3b = ggplot((df.med.g), aes(type,value)) + geom_boxplot(aes(fill=factor(type))) + scale_fill_manual(values=ct.colors)
p3b
```

Count LR pairs with strong autocrine cancer signaling across tumor types (median RC score > 40).

```{r}
count(df.med %>% filter(cc > 40))
count(df.med %>% filter(ss > 40))
count(df.med %>% filter(cs > 40))
count(df.med %>% filter(sc > 40))
```

## Top-5 C->C crosstalk pairs

```{r}
top5.lr = (df.med[order(-df.med$cc),])[1:5,'lr']
df.top5 = gather(df %>% filter(lr %in% top5.lr),'type','value', cc,sc,cs,ss,nn)
df.top5$type <- factor(df.top5$type, levels = c('cc','sc','cs','ss','nn'))
df.top5$t <- factor(df.top5$t, levels = c(c('MEDIAN'),setdiff(levels(df.top5$t),c('MEDIAN'))))
p3c = ggplot(df.top5, aes(x = "", y=value, fill = factor(type))) + geom_bar(width = 1, stat = "identity") + 
  theme(axis.line = element_blank(), axis.text = element_blank()) + scale_fill_manual(values=ct.colors) +
  labs(fill="class", x=NULL, y=NULL) + facet_grid(lr ~ t)
p3c
```

## Top-5 S->C crosstalk pairs

```{r}
top5.lr = (df.med[order(-df.med$sc),])[1:5,'lr']
df.top5 = gather(df %>% filter(lr %in% top5.lr),'type','value', cc,sc,cs,ss,nn)
df.top5$type <- factor(df.top5$type, levels = c('cc','sc','cs','ss','nn'))
df.top5$t <- factor(df.top5$t, levels = c(c('MEDIAN'),setdiff(levels(df.top5$t),c('MEDIAN'))))
p3d = ggplot(df.top5, aes(x = "", y=value, fill = factor(type))) + geom_bar(width = 1, stat = "identity") + 
  theme(axis.line = element_blank(), axis.text = element_blank()) + scale_fill_manual(values=ct.colors) +
  labs(fill="class", x=NULL, y=NULL) + facet_grid(lr ~ t)
p3d
```


# Top 10 stroma-stroma interactons

```{r}
topss.lr = (df.med[order(-df.med$ss),])[1:10,'lr']
df.topss = gather(df %>% filter(lr %in% topss.lr),'type','value', cc,sc,cs,ss,nn)
df.topss$type <- factor(df.topss$type, levels = c('cc','sc','cs','ss','nn'))
df.topss$t <- factor(df.topss$t, levels = c(c('MEDIAN'),setdiff(levels(df.topss$t),c('MEDIAN'))))
pss = ggplot(df.topss, aes(x = "", y=value, fill = factor(type))) + geom_bar(width = 1, stat = "identity") + 
  theme(axis.line = element_blank(), axis.text = element_blank()) + scale_fill_manual(values=ct.colors) +
  labs(fill="class", x=NULL, y=NULL) + facet_grid(lr ~ t) + theme(text = element_text(size = 7))
pss
```

## Expression Immune checkpoint ligands and receptors

```{r}
par(mfrow=c(2,3))
for(g in c('PDCD1','CD274', 'PDCD1LG2', 'CD86', 'CTLA4')) {
  ce = unlist(expr[g,paste0(names(tumor.colors),'_C')])
  se = unlist(expr[g,paste0(names(tumor.colors),'_S')])
  lim = c(0,max(ce,se))
  plot(ce,se,ylim=lim,xlim=lim,main=g,col=tumor.colors,cex=1.6)
  abline(0,1,lty='dashed')
}
plot(1,1)
legend('center',names(tumor.colors),col=tumor.colors,pch=21,cex=0.8)

```

## Tumor type specific LR interactions

```{r}
# require min 1 RPKM expression of both ligand and receptor
df.filt = df %>% filter(c_l > 1 |  s_l > 1 | n_l > 1) %>% filter(c_r > 1 |  s_r > 1 | n_r > 1)

par(mfrow=c(1,2))
pan.cc = spread(df.filt[,c('t','cc','lr')],'t','cc')
rownames(pan.cc) = pan.cc[,1]
pan.cc = pan.cc[,setdiff(colnames(pan.cc),c('lr','MEDIAN'))]
boxplot(pan.cc,horizontal = F,boxwex=0.6,cex=0.7,col=tumor.colors[colnames(pan.cc)],ylim=c(0,100),outcol='grey')

pan.sc = spread(df.filt[,c('t','sc','lr')],'t','sc')
rownames(pan.sc) = pan.sc[,1]
pan.sc = pan.sc[,setdiff(colnames(pan.sc),c('lr','MEDIAN'))]
boxplot(pan.sc,horizontal = F,boxwex=0.6,cex=0.7,col=tumor.colors[colnames(pan.sc)],ylim=c(0,100),outcol='grey')

# list top-2 for each cancer type
apply(pan.cc,2,function(x) row.names(pan.cc)[order(-x)[1:2]])
apply(pan.sc,2,function(x) row.names(pan.sc)[order(-x)[1:2]])
```

## stromal EGFR ligands in GBM

```{r}
par(mfrow=c(2,2))
egfr.ligands = c('AREG','BTC','EGF','HBEGF','SPINK1','TGFA','EFEMP1')
egfr.gbm = expr[egfr.ligands,c('GBM_C','GBM_S')]
plot(egfr.gbm[,1],egfr.gbm[,2],xlab='Cancer expr (log2)',ylab='Stroma expr (log2)',xlim=c(0,10),ylim=c(0,10),main='EGFR ligand expression in GBM')
text(egfr.gbm[,1],egfr.gbm[,2],egfr.ligands,pos=3,cex=0.8)
abline(0,1,lty='dashed')
egfr.ligands = c('AREG','BTC','EGF','HBEGF','SPINK1','TGFA','EFEMP1')
egfr.gbm = expr[egfr.ligands,c('LGG_C','LGG_S')]
plot(egfr.gbm[,1],egfr.gbm[,2],xlab='Cancer expr (log2)',ylab='Stroma expr (log2)',xlim=c(0,10),ylim=c(0,10),main='EGFR ligand expression in LGG')
text(egfr.gbm[,1],egfr.gbm[,2],egfr.ligands,pos=3,cex=0.8)
abline(0,1,lty='dashed')
```

# BRCA subtypes crosstalk analysis

## Enrichment of crosstalk in basal subtype

```{r}
df.brca = read.csv('data/BRCA_product_score.csv')
df.brca$r = sub('.+_',"\\1",df.brca$lr)
df.brca$l = sub('_.+',"\\1",df.brca$lr)
df.brca = df.brca %>% filter(t == 'IDC_Her2' | t == 'IDC_LumA' | t == 'IDC_LumB' | t == 'IDC_Basal') %>% droplevels()
# remove duplicated rows
df.brca = df.brca[!duplicated(df.brca),]
# 749 expressed LR pairs
#table(df.brca$t)

# Add pseudocounts to RC scores
df = df.brca
df$sum = (df$c_l+1)*(df$c_r+1)+(df$s_l+1)*(df$c_r+1)+(df$c_l+1)*(df$s_r+1)+(df$s_l+1)*(df$s_r+1)+(df$n_l+1)*(df$n_r+1)
df$cc = (df$c_l+1)*(df$c_r+1)/df$sum
df$sc = (df$s_l+1)*(df$c_r+1)/df$sum
df$cs = (df$c_l+1)*(df$s_r+1)/df$sum
df$ss = (df$s_l+1)*(df$s_r+1)/df$sum
df$nn = (df$n_l+1)*(df$n_r+1)/df$sum

cc.basal = (df %>% filter(t == 'IDC_Basal'))$cc
names(cc.basal) = (df %>% filter(t == 'IDC_Basal'))$lr
cc.others = apply(cbind(her2=(df %>% filter(t == 'IDC_Her2'))$cc,luma=(df %>% filter(t == 'IDC_LumA'))$cc,lumb=(df %>% filter(t == 'IDC_LumB'))$cc),1,mean)
cc.diff = log2((cc.basal)/(cc.others))
sc.basal = (df %>% filter(t == 'IDC_Basal'))$sc
names(sc.basal) = (df %>% filter(t == 'IDC_Basal'))$lr
sc.others = apply(cbind(her2=(df %>% filter(t == 'IDC_Her2'))$sc,her2=(df %>% filter(t == 'IDC_LumA'))$sc,her2=(df %>% filter(t == 'IDC_LumB'))$sc),1,mean)
sc.diff = log2((sc.basal)/(sc.others))

par(mfrow=c(1,2))
plot(cc.basal,cc.diff,pch=20,xlim=c(0.0,0.7),ylim=c(-5.5,5),cex=0.75,col=adjustcolor('#AAAAAA', alpha.f = 0.95))
abline(h=0,lty='dashed')
label.lr = function(key,col='red') {
  ids = grep(key,names(cc.basal))
  points(cc.basal[ids],cc.diff[ids],pch=20,xlim=c(0.0,0.7),ylim=c(-5.5,5),cex=0.8,col=adjustcolor(col, alpha.f = 0.99))
  text(cc.basal[ids],cc.diff[ids],names(cc.basal)[ids],cex=0.5,pos=4,col='black')
}
label.lr('_NOTCH1','green')
label.lr('_KIT','blue')
label.lr('_FZD7','red')
label.lr('_FZD2','orange')
label.lr('_RET','purple')
label.lr('_IL6ST','magenta')

plot(sc.basal,sc.diff,pch=20,xlim=c(0.0,0.7),ylim=c(-5.5,5),cex=0.75,col=adjustcolor('#AAAAAA', alpha.f = 0.95))
abline(h=0,lty='dashed')
label.lr = function(key,col='red') {
  ids = grep(key,names(sc.basal))
  points(sc.basal[ids],sc.diff[ids],pch=20,xlim=c(0.0,0.7),ylim=c(-5.5,5),cex=0.8,col=adjustcolor(col, alpha.f = 0.99))
  text(sc.basal[ids],sc.diff[ids],names(sc.basal)[ids],cex=0.5,pos=4,col='black')
}
label.lr('_NOTCH1','green')
label.lr('_KIT','blue')
label.lr('_FZD7','red')
label.lr('_FZD2','orange')
label.lr('_RET','purple')
label.lr('_IL6ST','magenta')
```

### RC scores for selected LR pairs in BRCA subtypes

```{r}
lrs = c('SFRP1_FZD2','WNT11_FZD7','WNT3_FZD7','CLEC11A_KIT','KITLG_KIT','MFAP2_NOTCH1','JAG1_NOTCH1','ARTN_RET','IL6_IL6ST')
df.basal.lrs = gather(df %>% filter(lr %in% lrs),'type','value', cc,sc,cs,ss,nn)
df.basal.lrs$type <- factor(df.basal.lrs$type, levels = c('cc','sc','cs','ss','nn'))
df.basal.lrs$lr <- factor(df.basal.lrs$lr, levels = lrs)
p3g = ggplot(df.basal.lrs, aes(x = "", y=value, fill = factor(type))) + geom_bar(width = 1, stat = "identity") + 
  theme(axis.line = element_blank(), axis.text = element_blank()) + scale_fill_manual(values=ct.colors) +
  labs(fill="class", x=NULL, y=NULL) + facet_grid(t ~ lr)
p3g
```

### Expression of selected ligands and receptors in BRCA subtypes

```{r}
subtype_colors = c(wes_palette(n=4, name="GrandBudapest1"),"#BBBBBB")

show = c('SFRP1','FZD7','NOTCH1','KIT','IL6ST','RET')
tmp = distinct(df.basal.lrs[df.basal.lrs$r %in% show,c('t','c_r','s_r','r')])
colnames(tmp) = c('t','c_l','s_l','l')
tmp = rbind(tmp,distinct(df.basal.lrs[df.basal.lrs$l %in% show,c('t','c_l','s_l','l')]))
colnames(tmp) = c('t','c_r','s_r','r')
tmp$r = factor(tmp$r, levels = show)

# add normal expression for each gene
for (rec in show) {
  ne = (df.basal.lrs %>% filter(r == rec) %>% select(n_r))[1,]
  if (is.na(ne)) {ne = (df.basal.lrs %>% filter(l == rec) %>% select(n_l))[1,]}
  tmp = rbind(tmp,data.frame(t='normal',c_r=as.numeric(ne),s_r=ne,r=rec))
}

p3ha = ggplot() + 
  geom_bar(data = tmp, aes(x=t, y=c_r, fill=t),stat = "identity") +
  scale_fill_manual(values=subtype_colors) + facet_grid(. ~ r) + scale_y_log10(lim =c(1, 250), breaks = c(10,100,1000))

p3hb = ggplot() + 
  geom_bar(data = tmp, aes(x=t, y=s_r, fill=t),stat = "identity") +
  scale_fill_manual(values=subtype_colors) + facet_grid(. ~ r) + scale_y_log10(lim =c(1, 250), breaks = c(10,100,1000))

pg = plot_grid(p3ha, p3hb, labels = c("a", "b"),ncol=1,align='v')
pg
```
